---
title: "Trait-based analysis with biomonitoR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{functional}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Trait-based analysis is a standard procedure in community ecology and represents a good complement to taxonomic information. Some efforts have also been made to include trait-based metrics into current biomonitoring practices. The `biomonitoR` package provides functions to manipulate trait data and to calculate functional metrics (e.g. richness, diversity). Some examples on how to deal with functional analysis in `biomonitoR` will be provided.

## Macroinvertebrates

For macroinvertebrates, `biomonitoR` provides by default the Tachet database (Tachet et al. 2010) available at [freshwaterecology.info](https://www.freshwaterecology.info/) website. Briefly, the Tachet database is a fuzzy-coded database with 21 categories and a total of 113 modalities. One of the first steps in trait-based analysis is to retrieve trait information for the taxa present in the taxonomic dataset. There are some time-consuming steps when retrieving traits for macroinvertebrates because of inconsistencies between taxonomic resolution of the taxonomic dataset and those of traits. `biomonitoR` accomplish this task with the function `traitScaling`.

```{r setup}
library(biomonitoR)

# import the example dataset in biomonitoR
# macro_ex includes taxa identified at species, genus and family level

data(macro_ex)

data.bio <- asBiomonitor(macro_ex)
data.agR <- aggregatoR(data.bio)


# use the traitScaling function

data.ts <- traitScaling( data.agR , taxLev = "Taxa" )

# we can use traitScaling at the desired taxonomic level
# for example setting taxLev = "Family"
# traitscaling does not allow taxonomic levels higher than families


```


`traitScaling` returns a `data.frame` reporting taxa names of the taxonomic dataset, the taxonomic level of each taxon of the taxonomic dataset, the name of the taxa whose traits have been associated to a taxon of the taxonomic dataset and the corresponding taxonomic level, the taxonomic distance between taxa of the taxonomic dataset and the one from trait dataset and traits. Taxonomic distance is calculated based on the taxonomic levels implemented in `biomonitoR`. There are some rules of aggregation that can be consulted on the help page of `traitScaling`.


```{r head traitScaling}
# data.ts has more rows because more entries have been found for one or more taxa.
# only the first two traits are showed to not fill the page with numbers

data.ts[ 1:5 , 1:7 ]


```


The `traitScaling` function links available traits to a taxon by looking at the taxonomic tree. Multiple assignements for the same taxon can thus be obtained.


```{r dim traitScaling}
# data.ts has more rows because more entries have been found for one or more taxa.

dim( macro_ex )
dim( data.ts )

# look at Beraeidae for which five taxa of the trait dataset have been associated
# to one taxon of the taxonomic dataset

data.ts[ data.ts$Taxa %in% "Beraeidae" , 1:5 ]


```

Taxonomic distance can be useful, for example when only exact matches are needed. The `filter_by_distance` option of the `traitScaling` function or simple subsets can be used.


```{r filter traitScaling}
traitScaling( data.agR , filter_by_distance = 0 )[, 1:5 ]

data.ts[ data.ts$Taxonomic_distance == 0 , 1:5 ]

```

What to do next? `biomonitoR` offers a more advanced subsetting tool with the function `manageTraits`.

```{r manage traitScaling}
# select the traits belonging to the nearest taxa that have a taxonomic level equal or finer than the target one

data.ts_plus <- manageTraits( data.ts , method = "nearest+" , traceB = TRUE )

# with the traceB option set to TRUE you can check the taxa removed because they did not meet the requirments
# Here Serratella ignita is at species level while the trait at genus level (Serratella)

data.ts_plus$taxa_excluded


```

When you are satisfied with the list of taxa you can finalize your trait dataset with two strategies. The function `traitsMean` allow the computation of the average traits for each Taxon. It is currently designed only for fuzzy data. For continuos data the default `aggregate` function can be used. Alternatively, taxa can be chosen randomly when multiple matches are available.


```{r mean traitScaling}
# take the average. If colB is set to NULL the default values will be used.
# Please consider that the default value is correct only for the default trait dataset!
# only the first 5 rows and columns will be showed

data.ts.av <- traitsMean( data.ts )
data.ts.av[ 1:5 , 1:5 ]

# alternatively a random selection can be done
# only the first 5 rows and  columns will be showed

sampleTraits( data.ts )[ 1:5 , 1:5 ]
```

Now we have obtained a trait dataset that can be used for calculating functional metrics.

### Functional indices

The calculation of functional dispersion, diversity, evenness, redundancy and richness are straightforward.

```{r functional indices }
colB <- c( 8, 7, 3, 9, 4, 3, 6, 2, 5, 3, 9, 8, 8, 5, 7, 5, 4, 4, 2, 3, 8 )

# functional dispersion
f_disp( data.agR , traitDB = data.ts.av , nbdim = 2 , type = "F" , colB = colB )

# functional diversity 
f_divs( data.agR , traitDB = data.ts.av , type = "F" , colB = colB )

# functional evenness
f_eve( data.agR , traitDB = data.ts.av , type = "F" , colB = colB )

# functional redundancy
f_red( data.agR , traitDB = data.ts.av , type = "F" , colB = colB )

# functional richness on 2 dimensions
f_rich( data.agR , traitDB = data.ts.av , nbdim = 2 , type = "F" , colB = colB )
```


### The traceB option

Some information on the calculation is provided by the traceB option. This option is implemented in several biomonitoR functions to have a feedback on the calculation and to check if functions beahved as intended. It is particularly important for functional indices.

```{r traceB }
frich <- f_rich( data.agR , traitDB = data.ts.av , nbdim = 2 , type = "F" , colB = colB , traceB = TRUE )

# look at the traits with NA values
head( frich$NA_detection )
```

### Community weighted mean and community specialization index


```{r cwm }

# community weighted mean and community specialization index
# only the first 5 columns are showed

cwm(x = data.agR, traitDB = data.ts.av, taxLev = "Taxa", trans = log1p )[ , 1:5 ]
csi(x = data.agR, traitDB = data.ts.av, taxLev = "Taxa", trans = log1p )[ , 1:5 ]

```


### Working with distance matrices
Distance matrices can be provided instead of the functional database for calculating functional indices. Building distance matrices assure more flexibility and control. `biomonitoR` deals only fuzzy and continuous traits only, but other types of traits are frequent in ecological applications.

```{r dist ma }
library( ade4 )

rownames( data.ts.av ) <- data.ts.av$Taxa
traits.prep <- prep.fuzzy( data.ts.av[ , -1 ], col.blocks = colB )

traits.dist <- ktab.list.df( list( traits.prep ) )
traits.dist <- dist.ktab( traits.dist , type = "F" )

f_rich( data.agR , traitDB = traits.dist , nbdim = 2 )
```

Functional richness and dispersion require the construction of a functional space from distance/dissimilarity matrices. `f_rich` and `f_disp` do it internally through when a trait dataset is provided, but more control can be obtained by providing a distance matrix. The fuction `selectPcoaAxes` is a useful tool to select the number of axes needed to build the functional space with a desired quality. Three methods for evaluating the qauality of the functional space are proposed. `selectPcoaAxes` also evaluates the performances of different correction algorhitms when negative eigenvalues are found during the construction of the functional space.


```{r pcoaAxes }
selectPcoaAxes( traits.dist , method = "cor" , tresh = 0.7)
selectPcoaAxes( traits.dist , method = "legendre" , tresh = 0.65)
selectPcoaAxes( traits.dist , method = "maire" , tresh = 0.01)
```


### Working at different taxonomic levels
One of the strength of `biomonitoR` is the ability to work at different taxonomic levels. The examples above were done at a mixed-taxonomic level (species, genus, family and order). If we wanto to look at the family level results,
we can repeat the analysis by specifying famly in both `traitScaling` and `indices`.


```{r family }
data_fam.ts <- traitScaling( data.agR , taxLev = "Family" )

data_fam.ts.av <- traitsMean( data_fam.ts )

# functional richness
f_rich( data.agR , traitDB = data_fam.ts.av , taxLev = "Family" , nbdim = 2 , type = "F" , colB = colB )

```



## Macrophytes
The same approach used for macroinvertebrates can be used for other groups. The following example is for macrophytes.


```{r macroph }
# load macrophyte data
data( oglio )

# transform to presence-absence data for simplicity
# ignore the warning
oglio.bin <- oglio
oglio.bin[ oglio.bin > 0 ] <- 1

# importing data in the biomonitoR format
oglio.asb <- asBiomonitor( oglio.bin , group = "mf" , FUN = bin )
oglio.agg <- aggregatoR( oglio.asb )

```




## REFERENCES
Schmidt-Kloiber, A., & Hering, D. (2015). An online tool that unifies, standardises and codifies more than 20,000 European freshwater organisms and their ecological preferences. Ecological Indicators, 53, 271-282.

Tachet H (2010) Invertébrés d’eau douce: Systématique, biologie, écologie, édition revue et augmentée edizione. CNRS Editions, Paris




