---
title: "Introduction to biomonitoR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
The biomonitoR package was primarily developed for dealing with taxonomic dataset and for calculating diversity, functional and biomonitoring indices at different taxonomic resolution. Although its focus is on freshwater macroinvertebrates, it actually works with multiple biotic groups.

## Import data into biomonitoR format
To import a taxonomic dataset in the `biomonitoR` format a `data.frame` with a column called "Taxa" where store taxa names and samples on the other columns is needed.

```{r setup}
library(biomonitoR)

# built-in example

data( macro_ex )

head( macro_ex )

```

The first step is to import taxonomic dataset with the function `asBiomonitor` that return an object of class `asb`. Basically this function  adds the taxonomic tree to the taxonomic data. The taxonomic tree is built from a reference database, specified in the option `group`. `biomonitoR` implements 3 reference database for macroinvertebrates, macrophytes and fish but a reference database can be added by the user.


```{r import}
# import macroinvertebrate data
data.bio <- asBiomonitor( macro_ex , group = "mi" )

data.bio
```


taxonomic dataset can be imported as presence-absence by levaragin on the `FUN` option. This option need to be thinked carefully. It is needed because `biomonitoR` accepts duplicated taxa names in the "Taxa" column. For abundance data `sum` is thus the natural choice for aggregating data. Case sensitive function for aggregating data could be used when decimal numbers are present (e.g. pecentages) and `biomonitoR` will remind this issue with a warning.

```{r import bin}
# import macroinvertebrate data
data.bio.bin <- asBiomonitor( macro_ex , group = "mi" , FUN = bin )

data.bio.bin
```

An important feature of the `biomonitoR` package is the spell checker, built from the reference database. By default `biomonitoR` remove taxa that are not recognized by the reference database and will provide suggestion for unrecognized names. The taxonomic dataset can thus be corrected by modifying the original file, directly in R or by using the spell checker. The spell checker is enabled by setting the option `correct_names` to to `TRUE` and it will run interactively. Every taxon needs to be present in the reference database, otherwise the user is forced to exit. This behaviour is important to assure consistency with further calculations. The option `traceB` will list suggestions and correction.


```{r spell}
# introduce an error into taxonomic data

macro_ex_wrong <- macro_ex
macro_ex_wrong$Taxa <- as.character( macro_ex_wrong$Taxa )
macro_ex_wrong[ macro_ex_wrong$Taxa %in% "Acentrella" , "Taxa" ] <- "Acentrela"

data.bio.wrong <- asBiomonitor( macro_ex_wrong , group = "mi" , traceB = TRUE )

# "Acentrela" has been removed and a suggestion is proposed
data.bio.wrong
```

`asBiomonitor` will create an object of class `asb` that is needed for perfroming further calculation in `biomonitoR`. It is possible to export both the taxonomic dataset and the suggestion/correction as data.frame objects using the function `as.data.frame`.

```{r as.data }
# export taxonomic dataset and suggestions as data.frame

macro_ex.df <- as.data.frame( data.bio.wrong )
macro_ex_correction.df <- as.data.frame( data.bio.wrong , object = 2 )
macro_ex_correction.df

```


The calculation of biomonitoring indices in `biomonitoR` can be done by creating an object with the function `aggregatoR`. This function aggregates data at various taxonomic levels.

```{r aggrega }
data.agg <- aggregatoR( data.bio )

```

It is possible to subset data prior to import them with the `aggregatoR` function. This is useful, for example, when the calculations need to be done with some Orders only.

```{r subset }
# select EPT Taxa (Ephemeroptera, Plecoptera and Trichoptera)

subset(data.bio, taxa = c( "Ephemeroptera" , "Plecoptera" , "Trichoptera" ) )

# select Trichoptera excluding the trichopteran family Hydropsychidae

tricho <- subset(data.bio, taxa = "Trichoptera" , exclude = "Hydropsychidae" )
tricho

# import tricho with the aggregatoR function
tricho.agg <- aggregatoR( tricho )

```


## Building your own reference database

Users can provide their own reference database, but they need to imported in the correct format. `biomonitoR` assists the user in creating their own reference database with the function `refFromTree`.

```{r refe }
# import data

data( Tree )

# Tree is a standard taxonomic tree that needs to be transformed into a reference database
# We can use the function refFromTree

ref_custom <- refFromTree(Tree)

ref_custom


```

The number of taxonomic levels of the reference database is 11 and their number and names cannot be modified. There is the hope to change this beahviour in a next version of `biomonitoR`.


